If the Null Fits, You Must Omit: Measurement Artifacts Undermine Group Polarization Replications

### 2025-10-24 GroupPolarizationStatmod/paper/intro-latest.tex
nearing complete Intro

But even if this measurement problem is fixed,
it is one of several severe problems in this kind of social science: Unstable
opinions, unmodeled input variance, and incommensurate experimental designs
all uniquely undermine any cumulative insight group polarization might
claim—problems known about and ignored for decades~\cite{Cartwright1973}. No
degree of scientific lawfulness can be established without evidence gathered
from rigorously modeled, commensurate experiments~\cite{Cartwright1999}.

@@

Group polarization research began in the 1960s, which grew rapidly with
funding from the National Institutes of Mental Health (NIMH), and widespread interest
in what was then a surprising new science of radicalization.
Dorian Cartwright (1973) reviewed the first NIMH grants and at the time said,
`` ''.

>>>>

But even if this measurement problem were fixed,
there are several problems in this kind of social science: Unstable
opinions, unmodeled input variance, and incommensurate experimental designs
all uniquely undermine any cumulative insight group polarization might
claim—problems known about and ignored for decades~\cite{Cartwright1973}. No
degree of scientific lawfulness can be established without evidence gathered
from rigorously modeled, commensurate experiments~\cite{Cartwright1999}.

@@

Group polarization research began in the 1960s, which grew rapidly with
focused government funding coupled with widespread
interest in what was then a surprising new science of
radicalization~\cite{Cartwright1971}. Dorian Cartwright (1973) reviewed the
first decade of group polarization research and found problems: he lamented
the ``disturbing fact'' that group polarization research has produced no
cumulative theoretical undersatnding of the effects of discussion on
group opinions~\cite{Cartwright1973}. Cartwright was somewhat dismayed, but
optimistically hoped that group polarizaiton research would eventually
produce ``an enduring\ldots more comprehensive paradigm'' \cite[p.
230]{Cartwright1973}.

Cartwright (1973) suggested that a ``vigorous search for alternative
formulations'' could help group polarization become more useful. More
researchers developed specific theories of group
polarization, with two becoming particularly influential: social comparisons
and persuasive arguments. The social comparisons hypothesis says group
polarization occurs because humans want to fit in, and so generally like to . Group polarization occurs, then, when a
group is composed mostly of extremists, with a sufficient number of people
with a weak opinion \cite{Myers1970; Myers1975}. Its main
“competitor” from the 1970s was the persuasive arguments theory, which
basically claimed that rhetoric was the strongest driver of group
polarization, not social comparisons \cite{Burnstein1973,Burnstein1975}.

Other notable hypotheses include the
self-categorization and social decisions scheme hypotheses.
Self-categorization augmented the social comparisons hypothesis with a
consideration of idiosyncratic beliefs people hold that are the result of
their own personal lived experience \cite{Abrams1990,Hogg1990}.
The social decisions hypothesis posits that contextual factors like
social network structure could be at least as important as the distribution
of opinions and . Noah Friedkin, a sociologist, notably published
the study with the least potential measurement artifacts for a rigorous study on group polarization, adopting the social
decisions scheme hypothesis, predicting that social connectivity within a
group could amplify group polarization \cite{Friedkin1999}.

\subsection{Measurement procedure}

In most group polarization experiments, opinions are measured on an
ordinal scale: Burnstein and Vinokur (1975) for example used a ten-point
scale from 1 to 10 representing “the lowest odds of success acceptable” to
try a certain behavior, where 1 means 10\%—Moscovici and Zavalloni (1969) gave participants a seven-point scale, from “strongly disagree” (-3) to “strongly agree” (+3) to report their opinions. In these and every other survey-based group polarization study I know, the reported opinions on the ordinal scale—with discrete numbered bins—were treated as if they were continuous-valued, or in other words as if they were the internal mental opinion value itself, whatever that might be neurobiologically. 
Liddell and Kruschke (2018) showed that in the general case this setup can
lead to false positives because ordinal scales cannot track changes in
extreme opinions beyond the ceiling of the measurement scale—this approach is
ubiquitous in social psychology and survey research, not at all limited to
group polarization.

\subsection{Study Corpus}

We selected ten journal articles with sixty total experimental conditions
across the ten.

\begin{table}[ht]
\caption{\textbf{Journal articles in our corpus.}}
\centering
\begin{tabular}{clcc}
\toprule
Authors (Year) & Title & Journal & Category \\
\midrule
  Moscovici \& Zavalloni (1969) & The Group as a Polarizer of Attitudes &
    Journal of Personality and Social Psychology & Politics \\
    Schkade, Sunstein, \& Hastie & When Deliberation Produces Extremism & Critical Review
                                 & Politics \\
Burnstein \& Vinokur (1973) & Testing Two Classes of Theories About Group Induced Shifts
In Individual Choice & Journal of Experimental Social Psychology & Persuasive Arguments \\

Burnstein \& Vinokur (1975) & What a person thinks upon learning he has chosen differently from others: Nice evidence for the persuasive-arguments explanation of choice shifts & Journal of Experimental Social Psychology & Persuasive Arguments \\
\bottomrule
\end{tabular}
\label{table:mr}
\end{table}
